Dutch, PPs, ʀ-pronouns

### 2019-12-22 paper.tex
zondagavond
- referenties aangepast
- eerste sectie bijna af
- stuctuur voor rest van verhaal gedaan

hoi

@@

This paper focuses on a single \tsc{r}-pronoun in a specific type of free relative construction, namely on the relative pronoun \tit{waarmee} `with what' in a free relative construction in which the predicates require different cases. The \tsc{r}-pronoun \tit{waarmee} `with what' is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with' when it is combined withn \tsc{r}-pronoun.

The construction I focus on is a free relative construction, in which the two predicates combine with different cases. I illustrate this in \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental PP. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative DP. The \tsc{r}-pronoun \tit{waarmee} `with what' is used here.

%%%%%%
Footnote
In this example \tit{mee} `with' is stranded and \tit{waar} `where' is moved to the left edge of the embedded clause. It is also possible for \tit{waar} `where' to bring \tit{mee} `with' along, as in \ref{ex:meealong} but it is regarded as less natural.

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:meealong}
Footnote
%%%%%%

@@

In this paper I show that distribution of \tit{waarmee} `with what' and \tit{met wat} `with what' in these free relative constructions gives us a unique insight into the internal structure of \tsc{r}-pronouns. In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a constituent, the \tsc{r}-pronoun surfaces. If the constituent is interrupted, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2006}.

@@

\section{\tsc{R}-pronouns as default}

@@

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waarmee} and \tit{met wat}}
	\begin{minipage}{0.27\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waarmee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✓             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✓             \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:distribution}

The goal of this section is to show that \tit{waarmee} is the default as instrumental relative pronoun. In order to do that, I discuss the distribution of \tsc{r}-pronouns and regular pronouns in more general. Dutch has the personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{het} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

@@

The example in \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. However, the inanimate personal pronoun \tit{het} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an ʀ-pronoun appears. This is illustrated in \ref{ex:preper}. \ref{erprep} shows that the \tsc{r}-pronoun obligatorily moves to the left of the pronoun.

@@

A picture similar to the one for inanimate personal pronouns appears with wh-inanimates.

object of verbs \tit{wat}.
object of preposition with \tit{wat}

then show they appear in the contexts below.

\tit{Waarmee} and not \tit{met wat} also appears in other contexts. \ref{ex:headless} shows matching headless free relatives, \ref{ex:headed} shows headed relatives, \ref{ex:wh} gives wh-questions. The use of \tit{met wat} is ungrammatical here, and \tit{waarmee} should be chosen.

\ex. Matching headless relatives\label{ex:headless}
\ag. Ik schilder waar jij ook mee schildert.\\
 I paint \tit{w-aa-r} you also with paint\\
 `I am painting with what you are painting with too.'
\bg. *Ik schilder met wat jij ook schildert.\\
 I paint with what you also paint\\
 `I am painting with what you are painting with too.'

\ex. Headed relatives\label{ex:headed}
\ag. Ik schilder met de kwast waar-mee jij ook schildert.\\
 I paint with the brush \tit{w-aa-r}-with you also paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

\ex. Wh-question\label{ex:wh}
\ag.  Waar-mee schilder jij?\\
 \tit{W-aa-r}-with paint you\\
 `What are you painting with?'
\bg. *Met wat schilder jij?\\
 With what paint you\\
 `What are you painting with?'

In sum, no inanimate pronominal elements in Dutch is able to act as the object of a preposition. It is always the \tsc{r}-pronoun that appears. What this shows is that ʀ-pronouns are not exceptions, but they are the default. There is something special about the examples in which \tit{met wat} is used. I suggest that the crucial difference lies in constituency. \tit{Waarmee} can only be used if the instrumental PP forms a proper constituent, i.e. a constituent to the exclusion of any other elements. --explain here how they all form a constituent. In the examples from xx to xx above this is the case.

There are actually more examples of \tit{wat} following \tit{met} in which \tit{met} and \tit{wat} do not form a constituent without containing any other elements. First, \tit{wat} is an indefinite that takes an NP complement, and only then it is combined with the \tit{met}. Second, \tit{wat} combines with \tit{voor} and then combines with an NP complement en then with \tit{met}. \tit{Met} and \tit{wat} are never a constituent to the exclusion of everything else.

\ex.
\ag. Ik wil graag thee met wat suiker.\\
 I want please tea with some sugar\\
 `I would like to have tea with some sugar.'
\bg. [Met [wat [voor [potloden]]] teken jij?\\
 with what for pencils draw you\\
 `What kind of pencils do you with?'

Let me now show how this applies to the examples with the mismatching relatives.

@@

\ex.
\ag. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'

Now let us return to the mismatching headed relatives. The two predicates I used in these are \tit{kopen} `to buy' and \tit{schilderen}  `to paint'. \tit{Kopen} takes an accusative DP as its object. \tit{Schilderen} can take an instrumental PP as its object.\footnote{Of course, \tit{schilderen} also possibly takes an accusative DP, but I am focussing on the instrumental here.}

\ex.
\ag. Ik koop het schilderij.\\
 I buy the painting\\
 `I am buying the painting.'
\bg. Ik schilder met een kwast.\\
 I paint with a brush\\
 `I am painting with a brush.'

\tit{Schildert} wants the instrumental PP, so the embedded clause looks like this:

\ex. \begin{forest} boom
[
    [
        [gekocht, roof]
    ]
    [
        [waar jij mee schildert, roof]
    ]
]
\end{forest}

Here \tit{waar} and \tit{mee} are both part of the embedded clause. At a certain point in the derivation they formed a proper constituent, i.e. they formed a constituent without any other elements.

\ex. \begin{forest} boom
[
    [
        [met, roof]
    ]
    [
        [wat jij hebt gekocht, roof]
    ]
]
\end{forest}

Then in the main clause \tit{gekocht} is merged, which selects for an accusative object. Somehow \tit{waarmee} can satisfy that requirement.

\tit{Schildert} wants the instrumental PP, so the embedded clause looks like this:

@@

\ex.
\a. *[[met] [wat]] $\rightarrow$ [waarmee]
\b. [met [wat X]]
\b. [met [wat [X]]]

@@

\section{internal structure of \tit{met wat} and \tit{waarmee}}

@@

--but first introduce a bit of theory--

\section{Taking \tit{waarmee} and \tit{met wat} apart}

>>>>

@@

This paper focuses on a single \tsc{r}-pronoun in a specific type of free relative construction. The \tsc{r}-pronoun I focus on is the relative pronoun \tit{waarmee} `with what', which is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with when it is combined with an \tsc{r}-pronoun.

The construction I focus on is a mismatching free relative. This is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I illustrate this in \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental PP. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative DP. The \tsc{r}-pronoun \tit{waarmee} `with what' is used here.
\footnote{In this example \tit{mee} `with' is stranded and \tit{waar} `where' is moved to the left edge of the embedded clause. It is also possible for \tit{waar} `where' to bring \tit{mee} `with' along, as in \ref{ex:meealong} but it is regarded as slightly less natural.

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:meealong}
\z.

\phantom{x}
}

@@

In this paper I show that distribution of \tit{waarmee} `with what' and \tit{met wat} `with what' in these free relative constructions gives us a unique insight into the internal structure of \tsc{r}-pronouns. In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a constituent, the \tsc{r}-pronoun surfaces. If the constituent is interrupted, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}.

@@

\section{The distribution between \tit{waarmee} and \tit{met wat}}

@@

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waarmee} and \tit{met wat}}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waarmee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:distribution}

In this section I first show that \tsc{r}-pronouns are the default complement of a preposition. Next, I illustrate that a necessary requirement for an \tsc{r}-pronoun is that is forms a proper constituent.

\subsection{\tsc{R}-pronouns as default}

The goal of this section is to show that \tit{waarmee} `with what' is the default as instrumental relative pronoun. This generalization is not new, it has already been made \citet{riemsdijk1978,koopman2000}. In order to show that \tit{waarmee} `with what' is the default, I discuss the distribution of \tsc{r}-pronouns and regular pronouns in more general. I start with the personal pronouns and then return to the wh-pronouns.

Dutch has the personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{het} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

@@

The example in \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. However, the inanimate personal pronoun \tit{het} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an ʀ-pronoun appears. This is illustrated in \ref{ex:preper}. \ref{ex:erprep} shows that the \tsc{r}-pronoun obligatorily moves to the left of the pronoun.

@@

The situation of the inanimate wh-pronouns resembles the inanimate personal pronouns. \tit{Wat} `what' can function as an object of a verb (see \ref{ex:wat}), but not as an object of a preposition \ref{ex:metwat}. In that case, the \tsc{r}-pronoun \tit{waarmee} `with what' is appears, as shown in \ref{ex:waarmee}.
\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes is only acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker questions the choice for the object hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 With what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar schilder jij mee?\\
 Where paint you with\\
 `What are you painting with?'\label{ex:waarmee}

\tit{Waarmee} `with what' and not \tit{met wat} `with what' does not only appear in wh-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental PP. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waarmee} `with what' is used.

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waar jij ook mee schildert.\\
 I paint with the brush where you also with paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waar jij ook mee schildert.\\
  I paint where you also with paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

Table \ref{tbl:inanimates} summarizes the distribution of inanimates in Dutch. Inanimate (personal and wh-)pronouns in Dutch can function as objects verbs, but they are ungrammatical as objects of prepositions. In these contexts, \tsc{r}-pronouns appear.

\begin{table}[ht]
	\center
	\caption {Inanimates in Dutch}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                        & pers. pronouns & wh-pronouns \\
  	\midrule
objects of verbs        & ət             & wat         \\
objects of prepositions & er             & waar        \\
\bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:inanimates}

The next section discusses the role of constituency in \tsc{r}-pronouns.

\subsection{\tit{Waarmee} is a constituent, \tit{met wat} is not}

Let me now return to the mismatching free relatives. I repeat the relevant grammatical examples in \ref{ex:grammatical}.

\ex.\label{ex:grammatical}
\ag. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:grammaticalwaar}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:grammaticalwat}

In this section I showed that ʀ-pronouns are expected in combinations with prepositions. This means that the use of \tit{waarmee} `with what' \ref{ex:grammaticalwaar} is not surprising. Something that is surprising is the use of \tit{met wat} `with what' in \ref{ex:grammaticalwat}, and this is the example something more needs to be said about. In the remainder of this section I argue that this `something more' is that the instrumental PP in \ref{ex:grammaticalwat} does not form a proper constituent, i.e. it is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns contain a PP that does form a proper constituent.

Below I repeat the examples with instrumental PPs I discuss so far in this paper.

\ex.
\ag. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar schilder jij mee?\\
 Where paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar jij ook mee schildert].\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar jij ook mee schildert].\\
 I paint where you also with paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

In each of these examples the instrumental PP forms a constituent at a certain point in the derivation. In \ref{ex:const1}, the PP forms a proper constituent in the surface order, as shown in \ref{ex:const1stage}. In \ref{ex:const2}, the PP forms a proper constituent before wh- and V2- movement, shown in \ref{ex:const2stage}.
The structure in \ref{ex:const3stage} represents a stage in the derivation of the embedded clauses in \ref{ex:const3} and \ref{ex:const4}.
\footnote{I assume that the antecedent in a free relative is a phonologically empty element, in line with cf. \cite{bresnan1978a,groos1981,himmelreich2017}.}
Again, in the stage, which comes before relative movement of the pronoun to the left periphery of the relative clause, the PP forms a proper constituent.

\ex.
\a. [[ik] [[schilder] [er-mee]]]\label{ex:const1stage}
\b. [[jij] [[schilder] [[waar] [mee]]]]\label{ex:const2stage}
\b. [[jij] [[ook] [[schilder] [[waar] [mee]]]]]\label{ex:const3stage}

There is additional evidence for the fact that \tit{waarmee} `with what' forms a constituent in the constructions above. In the example in \ref{ex:const2}-\ref{ex:const4} only \tit{waar} `where' was fronted, but \ref{ex:constwaarmee} shows that also the phrase containing \tit{mee} `with' can be moved. As wh-movement can only target constituents, it follows that \tit{waar} `where' and \tit{mee} `with' have to form a constituent.

\ex.\label{ex:constwaarmee}
\ag. Waarmee schilder jij?\\
 {where with} paint you\\
 `What are you painting with?'
\bg. Ik schilder [waarmee jij ook schildert].\\
 I paint {where with} you also paint\\
 `I am painting with what you are painting with too.'
\bg. Ik schilder met de kwast [waarmee jij ook schildert].\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'

\ref{ex:grammaticalwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both construction \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

\ex.\label{ex:moremetwat}
\ag. [Met [wat [voor [potloden]]] teken jij?\\
 with what for pencils draw you\\
 `What kind of pencils do you with?'\label{ex:watwasfur}
\bg. Ik wil graag thee [met [wat [suiker]]].\\
 I want please tea with some sugar\\
 `I would like to have tea with some sugar.'\label{ex:watindef}

Let me now show how this applies to the examples with the mismatching free relatives. The two predicates I used in the free relatives are \tit{kopen} `to buy' and \tit{schilderen}  `to paint'. \tit{Kopen} `to buy' takes an accusative DP as its object, illustrated in \ref{ex:kopen}. \tit{Schilderen} `to paint' can take an instrumental PP as its object., shown in \ref{ex:schilderen}.\footnote{Of course, \tit{schilderen} also optionally takes an accusative DP, but I am focussing on the instrumental here.}

\ex.
\ag. Ik koop het schilderij.\\
 I buy the painting\\
 `I am buying the painting.'\label{ex:kopen}
\bg. Ik schilder met een kwast.\\
 I paint with a brush\\
 `I am painting with a brush.'\label{ex:schilderen}

insert the zero thingy there

@@

\tit{Schildert} wants the instrumental PP, so the embedded clause looks like this. Here \tit{waar} and \tit{mee} are both part of the embedded clause. At a certain point in the derivation they formed a proper constituent, i.e. they formed a constituent without any other elements.

Then in the main clause \tit{gekocht} is merged, which selects for an accusative object. Somehow \tit{waarmee} can satisfy that requirement. \tit{Schildert} wants the instrumental PP, so the embedded clause looks like this:

@@

\ex.
\a. *[[met] [wat]] → [waarmee]
\b. [met [wat X]]
\b. [met [wat [X]]]

@@

\section{Connecting constituency and spellout}

Intro: why *\tit{met wat}? Not a filter, not a different paradigm.

\subsection{Korean negation and causative}

\subsection{nanostory of that}

\subsection{waarmee = parallel}

\section{A finer decomposition}

What I account for: w, a, why we see the locative

What I don't: similarity between \tit{met} and \tit{mee}, that preps do not change form

and i need to say something about the assumption i make for pronouns

@@
